---
title: "market"
format: pdf
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

## Strawberry-Market

Chenyi Wang, La Su, Zijing Zhao.

## Introduction

This project explores trends in the U.S. strawberry market using data from the USDA/NASS. The analysis focuses on three key variables: **market price**, **market value**, and **market supply**. Our goal is to understand how these factors have changed over time and how they relate to one another. We cleaned and merged multiple datasets to create a consistent dataset for exploratory data analysis (EDA).

## Analysis

## 1.price

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

straw_price <- read_csv("~/Desktop/market_price_received.csv", show_col_types = FALSE)

straw_price_clean <- straw_price %>%
  select(Year, State, Value) %>%
  rename(Price_CWT = Value) %>%
  mutate(Price_CWT = as.numeric(gsub("[^0-9.]", "", Price_CWT))) %>%
  filter(!is.na(Price_CWT)) %>%
  arrange(Year, State)
knitr::kable(
  straw_price_clean %>%
    head(10),
  caption = "Table 1. Preview of Cleaned USDA Strawberry Price Data (first 10 rows)",
  align = "c",
  booktabs = TRUE
)
```

Interpretation

The figure shows the average annual strawberry price in the U.S. from 1998 to 2024. Overall, prices have shown a steady upward trend, rising from around \$80-\$100 per CWT in the late 1990s to over \$140 per CWT in recent years. A sharp increase around 2010 marks a period of strong market demand and higher production costs, while the dip in 2020 likely reflects COVID-19 disruptions in labor and transportation. Prices then recovered quickly after 2021, indicating resilient consumer demand and supply adaptation. Overall, the data suggest that the U.S. strawberry market remains profitable and stable, though price volatility increases during years of market or environmental stress.

```{r}
library(ggplot2)
library(dplyr)

price_trend <- straw_price_clean %>%
  group_by(Year) %>%
  summarise(mean_price = mean(Price_CWT, na.rm = TRUE))

ggplot(price_trend, aes(x = Year, y = mean_price)) +
  geom_line(color = "#2E86C1", linewidth = 1.2) +
  geom_point(color = "#1B4F72", size = 2.2) +
  geom_smooth(method = "loess", se = FALSE, color = "#117A65", 
              linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Average U.S. Strawberry Price Received (1998–2024)",
    subtitle = "Measured in $ per CWT (100 lbs)",
    x = "Year",
    y = "Average Price ($/CWT)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.4)
  )
```

Interpretation

From 1998 to 2024, the average price received by U.S. strawberry growers for fresh-market sales shows a steady long-term increase with short-term fluctuations. Prices rose from about \$85 to nearly \$180 per CWT before 2010, reflecting growing consumer demand and higher production costs. Between 2011 and 2018, prices remained relatively stable at high levels. A sharp decline around 2020 likely resulted from COVID-19 disruptions and weather-related production issues. Since 2021, prices have rebounded and stabilized between \$140–150 per CWT, suggesting recovery but not a full return to earlier peaks. Overall, the trend reflects a resilient and competitive market, where technological improvements and global supply factors shape long-term price movements.

```{r}
library(ggplot2)
library(dplyr)

top_states <- c("CALIFORNIA", "FLORIDA", "OREGON", "NORTH CAROLINA")

ggplot(straw_price_clean %>% filter(State %in% top_states),
       aes(x = Year, y = Price_CWT, color = State)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Strawberry Prices by State (Fresh Market)",
    y = "Price Received ($/CWT)",
    x = "Year",
    color = "State"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

Interpretation

The figure compares fresh-market strawberry prices across major producing states. Overall, all states show an upward trend in prices over time, reflecting increasing production costs and steady consumer demand. Among them, California—the largest producer—maintains moderate but stable prices, while Florida and Oregon display larger year-to-year fluctuations, likely due to climate variability and smaller production scales. North Carolina shows a steady increase, gradually converging toward the national average. Overall, the data indicate that although regional conditions create short-term volatility, the general price pattern across states remains positive and consistent, supporting the resilience of the U.S. strawberry market.

**CONCLUSION:**

Overall, U.S. strawberry prices in the fresh market have shown a steady upward trend since the late 1990s, rising from about \$90 to \$140-150 per CWT. At the state level, Oregon consistently records higher prices than North Carolina, reflecting differences in production scale, climate, and market quality. Although prices have fluctuated over time, they have stabilized in recent years, indicating a mature and resilient market structure.

## 2.supply

```{r}
library(dplyr)
library(ggplot2)
library(readr)

supply_all <- read_csv("marketsupply_all_cleaned.csv", show_col_types = FALSE)
annual_total <- supply_all %>%
  group_by(Year) %>%
  summarise(total_value = sum(Value, na.rm = TRUE))

ggplot(annual_total, aes(x = Year, y = total_value)) +
  geom_line(linewidth = 1.2, color = "pink") +
  geom_point(size = 2, color = "pink") +
  labs(
    title = "U.S. Strawberry Market Supply (1998–2024)",
    subtitle = "Measured in total production (CWT, 100 lbs each)",
    x = "Year",
    y = "Total Production (CWT)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.title = element_text(face = "bold")
  )
```

Interpretation

The figure shows total U.S. strawberry production from 1998 to 2024. Overall, production increased steadily until around 2010, followed by a sharp fluctuation that likely reflects weather and market disruptions. After a temporary decline around 2020, production levels recovered and stabilized, suggesting improved efficiency and a resilient national supply chain.

```{r}
top_states <- supply_all %>%
  group_by(State) %>%
  summarise(avg_value = mean(Value, na.rm = TRUE)) %>%
  arrange(desc(avg_value)) %>%
  slice(1:5) %>%
  pull(State)

supply_top <- supply_all %>%
  filter(State %in% top_states)

ggplot(supply_top, aes(x = Year, y = Value, color = State)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2) +
  geom_smooth(se = FALSE, method = "loess", linetype = "dashed", color = "#7F7F7F") +
  labs(
    title = "Top 5 Strawberry-Producing States (1998–2024)",
    subtitle = "Measured in total production (CWT, 100 lbs each)",
    x = "Year",
    y = "Production (CWT)",
    color = "State"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

Interpretation

This figure illustrates production trends across the five leading strawberry-producing states in the U.S. California overwhelmingly dominates national strawberry output throughout the entire period, while Florida and North Carolina contribute at smaller but stable levels. Production from Oregon and the “Other States” category shows modest increases in later years, suggesting gradual diversification of strawberry cultivation beyond traditional growing regions.

**CONCLUSION:**

The combined dataset integrates USDA strawberry market supply records from 1998-2015 and 2016-2024, providing a continuous timeline for long-term trend analysis. Merging these datasets was necessary because each covered a different time span of the same variable — total state-level strawberry production. Combining them enabled a unified and consistent view of supply dynamics across decades.

As shown in Figure 1, total U.S. strawberry production has fluctuated notably over time, with clear peaks around 2009–2010 and 2017, followed by a moderate increase after 2020. These variations likely reflect both climatic impacts and market-driven factors influencing harvest volumes. Figure 2 highlights the dominance of California, which consistently accounts for the vast majority of national output. Meanwhile, Florida and North Carolina contribute smaller but stable shares, and minor increases among “Other States” suggest gradual diversification beyond traditional growing regions.

Overall, the merged dataset provides a comprehensive and coherent picture of long-term U.S. strawberry supply trends, facilitating robust interpretation and policy-relevant insights into the resilience and structure of the national market.

## 3.value

```{r, fig.width=7, fig.height=5, fig.align='center'}
library(dplyr)
library(ggplot2)
library(readr)

value_all <- read_csv("~/Desktop/marketvalue_clean.csv")

annual_value <- value_all %>%
  group_by(Year) %>%
  summarise(total_value = sum(Value, na.rm = TRUE))

p <- ggplot(annual_value, aes(x = Year, y = total_value)) +
  geom_line(linewidth = 1.3, color = "#FF0000") +
  geom_point(size = 2.3, color = "#FF0000") +
  geom_smooth(method = "loess", se = FALSE, color = "#0000FF", linetype = "dashed") +
  labs(
    title = "U.S. Strawberry Market Value (1998–2024)",
    subtitle = "Measured in total annual market value (USD)",
    x = "Year",
    y = "Total Value (USD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
print(p)

```

Interpretation

This line chart illustrates the total annual market value of U.S. strawberries from 1998 to 2024. The overall trend shows steady growth, with a sharp increase after 2010, a decline around 2017, and recovery in recent years. These fluctuations likely reflect market adjustments, consumer demand, and production costs. Overall, the pattern underscores the rising economic significance of strawberry production within the U.S. agricultural sector.

```{r, fig.width=7.5, fig.height=4.8, fig.align='center'}
library(dplyr)
library(ggplot2)
library(readr)

state_value <- value_all %>%
  group_by(State) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE)) %>%
  arrange(desc(mean_value))

ggplot(state_value, aes(x = reorder(State, mean_value), y = mean_value)) +
  geom_col(fill = "#AF7AC5", alpha = 0.9) +
  coord_flip(clip = "off") +
  geom_text(
    aes(label = round(mean_value / 1e6, 1)),
    hjust = -0.05,
    color = "gray20",
    size = 3.5
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Average Strawberry Market Value by State (1998–2024)",
    subtitle = "Measured in average annual market value (USD)",
    x = "State",
    y = "Average Value (USD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.subtitle = element_text(size = 9.5, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 40, 10, 20)
  )


```

Interpretation This horizontal bar chart compares the average market value of strawberry production across U.S. states between 1998 and 2024. California overwhelmingly dominates the market, contributing the majority of total value by a wide margin. Florida follows at a distant second, while the “Other States” category accounts for smaller but gradually increasing shares. Overall, the visualization highlights a strong regional concentration of economic activity in California, indicating that U.S. strawberry market value is heavily centralized within a single dominant state.

```{r}
california_value <- value_all %>%
  filter(State == "CALIFORNIA")

p <- ggplot(california_value, aes(x = Year, y = Value)) +
  geom_line(linewidth = 1.3, color = "#BB2649") +
  geom_point(size = 2.3, color = "#BB2649") +
  geom_smooth(method = "loess", se = FALSE, color = "blue", linetype = "dashed") +
  labs(
    title = "California Strawberry Market Value (1998–2024)",
    subtitle = "Measured in total annual market value (USD)",
    x = "Year",
    y = "Value (USD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )


```

Interpretation

This chart focuses on California, the leading state in U.S. strawberry production. From 1998 to 2015, California’s market value showed a steady and substantial increase, reflecting the state’s strong productivity and global demand. After 2016, the trend became more volatile, likely influenced by climate variability, labor costs, and export dynamics. Despite these fluctuations, California remains the core driver of the national strawberry economy, maintaining its dominant position over the entire study period.

**CONCLUSION:**

This strawberry market value analysis provides a comprehensive overview of economic trends in the U.S. strawberry industry from 1998 to 2024. Overall, total market value shows a strong upward trajectory, with notable fluctuations around 2017 and 2020, likely driven by climatic variation, market shifts, and production adjustments. California remains the dominant producer, accounting for the vast majority of national value, followed by Florida and a smaller contribution from other states. The steady rise in total value, alongside regional concentration, suggests that the U.S. strawberry industry has become increasingly valuable yet regionally dependent over time.

## **Summary**

Across all three dimensions—price, supply, and value—the U.S. strawberry market from 1998 to 2024 demonstrates both growth and volatility. Production and market value have generally increased, reflecting the expanding scale and economic significance of the industry. However, recurring fluctuations highlight the influence of external factors, such as climate variation, labor costs, and market demand. California continues to play a central and dominant role in both production and value, underscoring its structural importance in the national supply chain. Overall, the market appears strong and regionally concentrated, yet sensitive to external pressures, emphasizing the need for long-term sustainability and diversification strategies within the U.S. strawberry sector.

## Reference

Data Source: USDA National Agricultural Statistics Service (NASS).\
<https://www.nass.usda.gov/>
